Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-07-24 paper/paper.tex
Initial version of abstract and introduction without references.

\title{Surrogate model for aligned-spin binary neutron star waveforms using gaussian process regression}

@@

Construct 5d hierarchical surrogate model using reduced bases and GPR. Includes $\ell=2,3,4$, dynamical tides, 
and tapering fit to NR.

Provide framework for building high dimensional surrogate from NR.

@@

The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]]. From the nondetection of BNS events in aLIGO's first observing run (O1) the 90\% upper limit is still consistent with the upper end of this range [[O1UpperLimits]]. However, continued nondetection will begin to narrow down this range.

The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters are measured by matched-filtering of the GW data with waveform models, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors. [[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]]. Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]]. [[What about sky position? Does it matter?]]

The effective one body (EOB) formalism and numerical relativity (NR) simulations, on the other hand, have shown significantly smaller uncertainties for BNS systems in their respective ranges of validity than PN waveforms. For the inspiral, EOB waveforms have total phase uncertainties of $\sim 1$~rad as compared to PN waveforms with phase uncertainties of $\sim 10$~rad as determined by the variation from freedom within each formalism [[explain better. show your plot for this.]]. NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]].

Unfortunately, available codes for generating EOB waveforms for BNS systems can take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].

These quantities are measured with Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling.

Previous work on surrogate models has focused mainly on compressing the waveform parameter space into a linear combination of a small number of basis functions using either greedy methods [[cite]] or singular value decomposition [[cite]]. Significantly less work has been performed on optimally sampling waveforms so that the coefficients of these linear combinations of waveforms can be chosen through fitting or interpolation with the fewest number of waveform evaluations. Standard techniques such as tensor spline or Chebyshev interpolation require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper, we will develop methods to construct surrogate models from EOB and NR waveform codes that accurately reproduce waveforms from this high-dimensional parameter space, using as few waveform evaluations as possible. We will construct frequency-domain surrogates of the amplitude and phase.

--Overview of what was done in nonspinning BNS surrogate paper, and the improvements used for constructing aligned-spin surrogate.

--Advantages of using frequency domain surrogate (and why you didn't use it in last paper).

[[A notable effect that has not been included is the spin induced quadrupole moment of each NS.]]

We organize the paper as follows. In Section...

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\vec x$ and the hyperparameters in the Gaussian Process Regression model as $\vec\theta$.

@@

>>>>

\title{Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters. However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $>10^5$ waveforms that would be required with standard, grid-based interpolation methods. For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

%Finally, I will show how this same method can be used to optimally choose parameters for numerical relativity simulations to correct the the EOB surrogate during the last several orbits and merger.

@@

%The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]]. From the nondetection of BNS events in aLIGO's first observing run (O1) the 90\% upper limit is still consistent with the upper end of this range [[O1UpperLimits]]. However, continued nondetection will begin to narrow down this range.

The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]], now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]]. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]]. However, continued nondetection in runs now underway will narrow down this range.

%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters are measured with Bayesian parameter estimation that compares the GW data with waveform templates generated on-the-fly, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors. [[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]]. Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]]. [[What about sky position? Does it matter?]]

The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models which are truncated at the currently known 3.5PN order, lead to a uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz. This is approximately the same size as the contribution to the phase from tidal effects, and as a result the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]], and this also biases the recovered NS equation of state as well [[cite]]. Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveforms uncertainties of $\sim 10$~rad. NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\mathbf{x})$ given its waveform parameters $\mathbf{x}$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce the original training set of waveforms to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\mathbf{x})$ as a function of the waveform parameters $\mathbf{x}$. This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]]. Alternatively, one can use the empirical interpolation method [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \mathbf{x})$ at empirical nodes $T_j$. One then interpolates the waveform $h(T_j; \mathbf{x})$ as a function of $\mathbf{x}$ at each node $T_j$. This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], as well as for frequency-domain waveforms [[cite]].

Previous work on surrogate models has focused mainly on parameter spaces $\mathbf{x}$ with $<=3$~dimensions. This allows one to approximate the coefficients $c_i(\mathbf{x})$ or waveform values $h(T_j; \mathbf{x})$ as functions of $\mathbf{x}$ using simple fitting functions [[cite]], or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]]. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR) that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. A similar approach has recently been used to optimally sample waveforms for a 2-dimensional parameter space using GPR [[DoctorEtAl]]. GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation [[ChrisMoore2papers?]].

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform a using slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]]. Whereas in [[Ben]] we constructed a time-domain surrogate for the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.

We organize the paper as follows. In Section II we describe the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space. In section III, we describe the details of building the surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in section IV. Then, finally discuss future improvements and applications in Section V.

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\mathbf{x}$ and the hyperparameters in the Gaussian Process Regression model as $\mathbf{\theta}$.

@@

\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]]. The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}
